---
title: "Effect of Vitamin C on Tooth Growth in Guinea Pigs"
author: "Karthik Arumugham"
date: "29 April 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path='Figs/', warning=FALSE, message=FALSE)
```

##Overview
The purpose of this data analysis is to analyze the ToothGrowth dataset by formulating several hypothesis for comparing the tooth growth of guinea pigs by supplement and dose.

```{r, echo=FALSE}
library(datasets); library(ggplot2)
data(ToothGrowth)
str(ToothGrowth)
```

The dataset has 60 observations, length of odontoblasts (cells responsible for tooth growth) in each of 10 guinea pigs at three dose levels of Vitamin C (0.5, 1 and 2 mg) with two delivery methods (orange juice or ascorbic acid).

```{r}
summary(ToothGrowth)
```


##Exploratory Analysis

1. **Plot 1**: The tooth lengths seem to be normally distributed.
2. **Plot 2**: There is a  large variation and clear seperation in tooth growth for all dose levels of ascorbic acid. But there appears to be clear seperation only for 0.5 and 1 mg/day dose levels of orange juice.

##Hypothesis Testing

**Hypothesis 1**:

- $H_0$: Mean tooth growth from orange juice is same as ascorbic acid.
- $H_1$: Mean tooth growth from orange juice is higher.

```{r}
h1<-t.test(len ~ supp, ToothGrowth, paired = F, var.equal = T)
h1$conf.int; h1$p.value
```
The p-value >0.05 and C.I. includes 0. The $H_0$ cannot be rejected. So orange juice & ascorbic acid seem to deliver the same tooth growth without factoring dose levels.

**Hypothesis 2**:

- $H_0$: Mean tooth growth from orange juice is same as ascorbic acid for the dose level of 0.5 mg/day.
- $H_1$: Mean tooth growth from orange juice is higher for the dose level of 0.5 mg/day.

```{r}
h2<-t.test(len ~ supp, data = subset(ToothGrowth, dose == 0.5), paired = F, var.equal = T)
h2$conf.int; h2$p.value
```
The p-value is <0.05. The C.I. does not include 0. The $H_0$ is rejected. So 0.5 mg/day dosage of orange juice seems to deliver more tooth growth than ascorbic acid.

**Hypothesis 3**:

- $H_0$: Mean tooth growth from orange juice is same as ascorbic acid for the dose level of 1 mg/day.
- $H_1$: Mean tooth growth from orange juice is higher for the dose level of 1 mg/day.

```{r}
h3<-t.test(len ~ supp, data = subset(ToothGrowth, dose == 1), paired = F, var.equal = T)
h3$conf.int; h3$p.value
```
The p-value is <0.05 and the C.I. does not include 0. The $H_0$ is rejected. So 1 mg/day dosage of orange juice seems to deliver more tooth growth than ascorbic acid.

**Hypothesis 4**:

- $H_0$: Mean tooth growth from orange juice is same as ascorbic acid for the dose level of 2 mg/day.
- $H_1$: Mean tooth growth from orange juice is higher for the dose level of 2 mg/day.

```{r}
h4<-t.test(len ~ supp, data = subset(ToothGrowth, dose == 2), paired = F, var.equal = T)
h4$conf.int; h4$p.value
```
The p-value is >0.05 and the C.I. does include 0. $H_0$ cannot be rejected. So 2 mg/day dosage of orange juice seems to deliver same tooth growth as ascorbic acid.

**Hypothesis 5**:

- $H_0$: Mean tooth growth from ascorbic acid (2 mg/day) is same as orange juice (1 mg/day).
- $H_1$: Mean tooth growth from ascorbic acid (2 mg/day) is higher than orange juice (1 mg/day).

```{r}
temp1 <- rbind(subset(ToothGrowth, (dose == 1) & (supp =="OJ")),
               subset(ToothGrowth, (dose == 2) & (supp =="VC")))
h5<-t.test(len ~ supp, temp1, paired = F, var.equal = T)
h5$conf.int; h5$p.value
```
The p-value is >0.05 and the C.I. does include 0. The $H_0$ cannot be rejected. 2 mg/day of ascorbic acid seems to deliver same tooth growth as 1 mg/day of orange juice.

**Hypothesis 6**:

- $H_0$: Mean tooth growth from orange juice (2 mg/day) is same as orange juice (1 mg/day).
- $H_1$: Mean tooth growth from orange juice (2 mg/day) is higher than orange juice (1 mg/day).

```{r}
temp2 <- rbind(subset(ToothGrowth, (dose == 1) & (supp =="OJ")),
               subset(ToothGrowth, (dose == 2) & (supp =="OJ")))
h6<-t.test(len ~ dose, temp2, paired = F, var.equal = T)
h6$conf.int; h6$p.value
```
The p-value is <0.05 and the C.I. does not include 0. $H_0$ is rejected. So 2 mg/day of orange juice seems to deliver higher tooth growth than 1 mg/day.

##Conclusion
Increase in the dose levels seem to increase the tooth growth. Orange juice is more effective than ascorbic acid for tooth growth when the dosage is 0.5 and 1 mg/day. Both types of supplements are equally as effective when the dose is 2 mg/day. I've assumed there's a common variance in the guinea pigs population (var.equal=TRUE) and that they've used 6 different settings of 10 guinea pigs for the experiment (paired=FALSE) so as not to get biased by an earlier test.

##Appendix

**Plot 1: Distribution of Tooth Length**

```{r, echo=FALSE}
hist(ToothGrowth$len, breaks=10, xlab = "Tooth Length", main="")
```

**Plot 2: Tooth Length vs Supplement and Dose Levels**

```{r, echo=FALSE}
ggplot(ToothGrowth, aes(x=factor(dose), y=len)) + facet_grid(.~supp) + geom_boxplot(aes(fill = supp), show.legend = FALSE) + labs(x="Dose (mg/day)",y="Tooth Length")
```

**Plot 3**:
```{r}
qqnorm(ToothGrowth$len); qqline(ToothGrowth$len, col = 2)
```

