---
title: Is Exponential Distribution normal?
author: "Karthik Arumugham"
date: "29 April 2016"
output: pdf_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(fig.path='Figs/', warning=FALSE, message=FALSE)
```

##Overview
The purpose of this data analysis is to investigate the exponential distribution and compare it with the properties of a normal distribution.

##Exponential distribution simulation
The simulation generates a distribution of averages of 40 exponentials (lambda set to 0.2) over 1000 simulations.
```{r}
set.seed(1)
lambda <- 0.2; n <- 40; s <- 1000
sim <- replicate(s, rexp(n, lambda))
meanE <- apply(sim, 2, mean)
```
###Comparison of sample mean and theoretical mean of the distribution
```{r}
meanS <- mean(meanE); meanS
meanT <- 1/lambda; meanT
```
The analytics mean is `r round(meanS,3)`, the theoretical mean is `r round(meanT,3)`. The center of distribution of averages of 40 exponentials is very close to the theoretical center of the distribution. The **Plot 1** shows the distribution with both the means.

###Comparison of sample variance and theoretical variance of the distribution
```{r}
varS <- (sd(meanE))^2; varS
varT <- ((1/lambda)*(1/sqrt(n)))^2; varT
```
Standard Deviation of the distribution is `r round(sqrt(varS),3)` with the theoretical SD of `r round(sqrt(varT),3)`.

###Check if distribution is approximately normal

From **Plot 2**, the distribution of averages of 40 exponentials appears to be very close to a normal distribution. Also the C.I. for 1/lamda accomodates the simulated mean. $H_0$ that difference between both means is 0 is failed to be rejected.

```{r}
mean(meanS) + c(-1, 1) * 1.96 * sqrt(varS)/sqrt(n)
```

From **Plot 3**, the Q-Q plot shows that normality is probably a reasonably good approximation. But the Shapiro-Wilk normality test rejects the $H_0$ that 
"the samples come from a normal distribution" against the $H_1$ that "the samples do not come from a normal distribution"
```{r}
shapiro.test(meanE)
```

##Appendix

**Plot 1**: Histogram of means of exponential distribution
```{r}
hist(meanE, xlab = "mean", main = "")
abline(v = meanS, col = "blue")
abline(v = meanT, col = "grey")
```

**Plot 2**: Exponential vs Normal Distribution
```{r}
xfit <- seq(min(meanE), max(meanE), length=100)
yfit <- dnorm(xfit, mean=meanT, sd=sqrt(varT))
hist(meanE,breaks=n,prob=T,xlab = "Means",main="",ylab="Frequency"); lines(xfit, yfit, pch=25, lty=5)
```

**Plot 3**:
```{r}
qqnorm(meanE); qqline(meanE, col = 2)
```

